---
title: "A model for underdispersed counts, or looking for data!"
author: "Noam Ross"
date: "February 13, 2015"
output: html_document
runtime: shiny
---

```{r setup, echo = FALSE, message=FALSE}
library(devtools)
#install.packages('ggplot2')
#install.packages('gridExtra')
#install_github('noamross/noamtools')
#install_github('noamross/cmp')
library(cmp)
library(ggplot2)
library(gridExtra)
library(noamtools)
```

The Conway-Maxwell Poisson (CMP) distibution is a distribution that can be used
to model discrete count data.  It has the form:

$$P(X = x) = \frac{\lambda^x}{(x!)^\nu} \frac{1}{Z(\lambda, \nu)}$$

where

$$Z(\lambda, \nu) = \sum_{j=0}^\infty \frac{\lambda^j}{(j!)^\nu}$$

Like the Poisson distribution, $\lambda$ is an encounter rate. Unlike the Poisson
distribution, the $\nu$ term modifies it.  When $nu < 1$, the distribution is *overdispersed*; its tail extends
out further, and its variance is greater than its mean. When $nu > 1$, the distribution
is *underdispersed* - its tail is shortened, and its variance is less than
its mean.  If $\nu = 1$, the CMP distribution is a Poisson distribution.  If $\nu = \infty$, the CMP distribution
is the same as a Bernoulli distribution.[^1]

[^1]: All this is in *Shmueli, G., Minka, T. P., Kadane, J. B., Borle, S. and Boatwright, P. (2005) A
useful distribution for fitting discrete data: Revival of the 
Conway-Maxwell-Poisson distribution.  J. Royal Statist. Soc. vol. 54, pp. 127-142, 2005.*

***I'm looking for examples of data
sets to fit with this distribution.*** I am testing a [draft R package](https://github.com/noamross/cmp) and need more ecological examples. Some ecological processes that might produce underdispersed count data are:

-   ***Density dependence/competition*:** If you plants or benthic organisms
    compete for space, counts of those organisms (in say, quadrats), can be under-dispersed
    because even if they arrive randomly at a location, thinning makes
    locations with population densities rate.
    
-   ***Disease mortality*:** If parasites in an organism cause mortality, counts
    of parasites in organisms may be underdispersed, because organisms with lots
    of parasites will have died.
    
If you know of/have an ecological dataset that may have underdispersed counts, for
these reasons or others let me know! ([\@noamross](http://twitter.com/noamross), [noam.ross\@gmail.com](mailto:noam.ross@gmail.com)).

## Look! Shiny buttons!

To give you an idea of how the CMP distribution works, here's an interactive chart.
Adjust the sliders to modify the $\lambda$ and $\nu$ values.  Checking the box
will also show a Poisson distribution with an equivalent $\lambda$ value. The
red vertical line shows the mean of the CMP distribution.

You'll note that when $\lambda$ is big and $\nu$ is small, calculating the probability
takes a little bit of time.

```{r inputs, echo = FALSE, message=FALSE}
inputPanel(
  sliderInput("lambda", label = "Lambda:",
              min = 0, max = 2.5, value = 1, step = 0.1),
  sliderInput("nu", label = "Nu:",
              min = 0.1, max = 5, value = 1, step = 0.01),
  checkboxInput("pois", "Show Poisson Distribution?", value=FALSE)
)
```

```{r makeplot, echo = FALSE, message = FALSE}
renderPlot({
  max_x = max(qcmp(0.99,input$lambda,input$nu),
              qpois(0.99, input$lambda))
  data = data.frame(Value = 0:max_x,
                    Probability = dcmp(0:max_x, lambda=input$lambda,
                                       nu=input$nu),
                    Distribution="CMP")
  if(input$pois) {
    data = rbind(data, data.frame(Value=0:max_x, 
                                  Probability = dpois(0:max_x,
                                                      lambda=input$lambda),
                                  Distribution="Poisson"
                                  ))
  }
  meanval = cmp_mean(lambda = input$lambda, nu = input$nu)
  cmpplot = ggplot(data, aes(x=Value, y=Probability,
                             fill=Distribution, linetype=Distribution,
                             lwd=Distribution)) 
  if(max_x <= 50) {
    cmpplot = cmpplot + geom_bar(stat="identity", position="identity", alpha=0.5) + scale_x_discrete(limits=0:max_x)
    
  } else {
    cmpplot = cmpplot + geom_area(alpha=0.5)
  }

    
  cmpplot = cmpplot +  theme_nr +
    geom_vline(xintercept=meanval, col="red", lwd = 2) +
    annotation_custom(grob=textGrob(substitute(paste(lambda, ": ", LL, "\n",
                                                     nu, ": ", NN, ", ",
                                                     E(X), ": ", MM, ", "),
                                               list(LL = input$lambda,
                                                    NN = input$nu,
                                                    MM = signif(meanval,3))),
                                    x = 0.8, y= 0.8))
  
  if(input$pois) {
    cmpplot = cmpplot +
      scale_linetype_manual(values=c(0,2))
  }

  cmpplot
}, width=800, height=400)  
```
