---
title: "Exploring VegBank Data Counts"
author: "Noam Ross"
date: "2/18/2015"
output: html_document
---

The kind folks at [VegBank](vegbank.org), via the the help of [Matt Jones](http://twitter.com/metamattj)
provided me with a complete dump of the VegBank database to explore.

The [VegBank database structure](http://vegbank.org/vegdocs/design/erd/vegbank_erd.pdf) is complex.

First, I need to create the database out of the ~10 GB dump provided.

PostgreSQL is required for this (and the subsequent operations that I do in R)

```{r createdb, engine='bash', eval=FALSE}
createdb -T template0 vegbank #create an empty database
psql vegbank < infile
```

OK, now that we have that, let's get some basic info on this database:

```{r db_info}
library(DBI)
library(RPostgreSQL)
veg_db = dbConnect(dbDriver("PostgreSQL"), dbname="vegbank")
dbListTables(veg_db)
```

```{r}
dbDisconnect(veg_db) #disconnect, because now we'll manage this with dplyr
rm(veg_db)
```

```{r}
library(dplyr)
veg_db = src_postgres("vegbank")
stemCount = tbl(veg_db, "stemCount")
```



