---
title: "Underdispersion in Macroparasite Models"
author:
- affiliation: UCD
  email: nmross\@ucdavis.edu
  footnote: Corresponding author
  name: Noam Ross
date: "August 17, 2015"
output:
  pdf_document:
    dev: png
    fig_caption: yes
    fig_height: 6
    fig_width: 11
    keep_tex: yes
    template: elsarticle.latex
  html_document:
    keep_md: yes
csl: ecology-letters.csl
address:
- address: |
    Department of Environmental Science and Policy
    University of California-Davis
    1 Shields Avenue Davis, CA 95616 USA
  code: UCD
documentclass: elsarticle
layout: review,12pt
linenumbers: yes
bibliography: references.bib
abstract: In multi-infection, or macroparasite, models of disease, distributions of
  infections among individuals are critical to host-parasite dynamics, Most analysis
  has focused on the form and stability of equilibrium distributions.  Here I use
  a classic macroparasite model framework, implemented as both ordinary differential
  equations and as a stochastic individual-based model, to examine the transient dynamics
  of such infections. The continuous form can produce underdispersed distributions
  of infections, even in the absence of density dependence. This underdispersion is
  weakened in the stochastic version of the model. In all cases, the Conway-Maxwell-Poisson
  distribution [@Conway1962] better describes the evolving distribution than the widely
  used negative binomial distribution. I also introduce an R package, **cmp**, for
  fitting this distribution to data.
---

```{r get_chunks, echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE}
library(knitr)
opts_knit$set(eval.after = "fig.cap")
read_chunk('setup.R')
read_chunk('ode_sims.R')
read_chunk('micro_sims.R')
read_chunk('micro_process.R')
read_chunk('ode_plots.R') 
read_chunk('micro_plots.R') 
opts_chunk$set(echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, dpi=144)
```

# Introduction

Aggregation of parasites within hosts is a very common phenomenon [@Shaw1995],
so much so that it has been described as a "law" of parasite ecology 
[@Poulin2007a]. Aggregation describes the phenomenon of a small number of hosts 
housing a larger fraction of the parasite population than would be expected by 
chance. It is characterized by the distribution of parasite loads having a 
variation larger than its mean, or overdispersion. This arises from 
heterogeneities such as individual variation in host susceptibility and 
exposure to parasites [@Keymer1979; @Anderson1982], and "clumping" of parasites 
that colonize hosts together [@Pugliese1998].

In rare cases, parasite loads may be underdispersed, with parasites more evenly 
distributed among hosts. Forces driving underdispersion include 
density-dependent parasite mortality and nonlinear parasite-induced host
mortality [@Barbour2000]. Underdispersed parasite distributions have been found
in oxyuroid parasites of cockroaches [@MullerGraf2001; @Zervos1988], in which 
sexual competition appears to induce density dependence. There are also a 
variety of examples in fish populations: @Burn1980 found nearly all Deretrema 
parasites in flashlight fish (*Anomalops katopteron*) were in groups of 2. In 
other cases, parasite numbers greater than one are rare [@Donnelly1994; 
@Adlard1994; @Kennedy1979]. Often, this density dependence is driven by limited 
space in the host [@Uebelacker1978]. The rarity of underdispersal suggests that
the influence of density-dependence is less than than that of individual
variation in most populations, despite the fact that density-dependence does
occur in many parasite populations [@Poulin2007].

Most models of host-macroparasite dynamics are variants of that of @Anderson1978, who used
an infinite-dimensional system of ordinary differential equations (ODE) to
represent a host population partitioned into indefinite number of classes
according to their parasite load. The infinite-dimensional system was reduced
to a two-dimensional system by assuming parasites were distributed by fixed
Poisson or negative-binomial distributions. Later, it was shown that these
models are equivalent in terms of stability [@Kretzschmar1993a], though the
exact form of the equilibrium distribution varies some from the
negative-binomial [@Barbour2000]. @Kretzschmar1993b showed that the
variance-mean ratio is important to system stability; it must increase with the
mean for there to be an internal host-parasite equilibrium. This suggests why
under-dispersed populations are rare - they lead to extinction of the parasite
population.

These analyses have focused on the equilibrium behavior of host-parasite models.
Even the absence of density dependence or individual variation, though
distributions of macroparasites evolve over time [@Adler1992]. The transient
dynamics of host-parasite populations are of interest in understanding and
managing emerging diseases, outbreaks, and cases of potential extinction.
Spore-driven fungal diseases, such chytrid fungus, sudden oak death,
and white nose syndrome, meet these criteria.  These emerging disease are 
capable of causing local extinctions in host populations [@Briggs2010;
@Fisher2012; @James2015; @Langwig2015]. While not traditionally considered
macroparasites, the dynamics of these diseases are similar; disease loads in
hosts are driven by repeated infections rather than internal reproduction,
and host mortality is load-dependent [@Briggs2010; @Langwig2015].

Here I examine the dynamics the distribution of parasites among hosts in the
classic infinite-dimensional model during the emergence of a disease that
suppresses or host populations. I compare these dynamics to an equivalent
discrete, stochastic, individual-based model. For both, I examine the relative
utility of the negative binomial and the Conway-Maxwell-Poisson distributions
in describing these distributions as they change over the course of an epidemic.

# Methods

## Models

I use two versions of a model closely related to the macroparasite model
introduced by @Anderson1978 and examined in much
of the subsequent literature [@Adler1992; @Kretzschmar1993a; @Kretzschmar1993b;
@Barbour2000]. The primary difference is the separation of birth
and death processes, with host density dependence only applying to the birth
process. In the model, the host population $N$ is subdivided into classes with 
infection level $i$, where $i$ is discrete and ranges from 1 to $\infty$.  New, uninfected hosts recruit into the population at a density-dependent rate
$r (1 - \sum N_i / K)$. Hosts gain infections at rate $\Lambda$,
which is the sum of total infections among all hosts $(\sum i N_i)$ times a
contact rate $\lambda$. Hosts recover from infections at a per-infection rate
$\mu$, and die at a background rate $d$ plus a per-infection rate $\alpha i^p$,
where the power term $p$ represents non-linearity in mortality. The full model
is described by:

$$\begin{aligned}
  N_0 &= r \sum_{i = 0}^\infty N_i \left(1 - \frac{\sum_{i = 0}^\infty 
N_i}{K}\right) - (d + \Lambda) N_0 \\
  N_i &= \Lambda N_{i-1} - \Lambda N_i - \alpha i^p - \mu i N_i + \mu (i + 1) 
N_{i+1} \\
  \Lambda &= \lambda \sum_{i = 1}^\infty i N_i
\end{aligned}$$

`r options(scipen = 1, digits = 2)`

For purposes of simulation, I truncate the number of infection classes at
$i=100$. For all parameterizations shown here, the value of $N_{100}$ never
differed from zero within machine tolerance (approximately $`r .Machine$double.eps`$).

The second version is the equivalent individual-based model (IBM), which includes demographic 
stochasticity.  In the IBM, the host population is represented as a discrete set of individuals, each with a
discrete number of infections. Stochastic rate equations determine the rate at which
each individual $j$ with infections $i$ may reproduce, die, acquire a new infection,  or recover from infections. The rates
$r$ are governed by the same parameters as in the ODE model above. The full set
of rate equations are:

$$\begin{aligned}
r_{j, \, birth} &= \max \left\{ r(1-N/K), \, 0 \right\} \\
r_{j, \, death} &= \alpha i_j^p + d \\
r_{j, \, infection} &= \Lambda \\
r_{j, \, recovery} &= \mu i_j \\
\Lambda = \lambda \sum_{j = 1}^N i_j \\
\end{aligned}$$

Individual states are updated in continuous time, using Gillespie's [-@Gillespie1976] stochastic simulation algorithm (SSA).

```{r loadpkgs, cache=FALSE, message=FALSE}
```

```{r setparms, cache=FALSE}
```

Table 1 shows the model parameters used in all cases in this paper.

| Parameter | Symbol | Value |
|-----------|--------|-------|
| birth rate | $r$  | `r paste(signif(r_vals, 1), collapse = ", ")` |
| carrying capacity | $K$ | `r format(baseparms$K)` |
| intrinsic host mortality rate | $d$ | `r format(baseparms$d)` |
| intrinsic parasite mortality rate | $\mu$ | `r format(baseparms$mu)` |
| additional host mortality per infection | $\alpha$ | `r format(baseparms$alpha)` |
| power of host mortality per infection | $p$ | `r paste(signif(alpha_power_vals, 1), collapse = ", ")` |
| contact rate | $\lambda$ | `r format(baseparms$lambda, scientific = FALSE)` |
| initial host population | $N_0$ | `r format(baseparms$N0)` |
| initial parasite population | $P_0$ | `r format(baseparms$P0)` |
| time period | $T$ | `r format(baseparms$time_max)` |

  Table 1: Parameters for the host-parasite models

I examine transient dynamics by simulation across a range of growth rates $r$
and with both linear $(p = 1)$ and nonlinear $(p = 2)$ parasite-induced mortality.
The parameters, shown in Table 1, all represent conditions under which the
parasite can invade the host population and the ODE system reaches an equilibrium
with a suppressed host population.

The ODE simulations were performed in R [@RCORE], using the deSolve
package [@deSolve]. The IBM simulations were performed using a custom package [TODO].
Code to reproduce the simulations is available at TODO.

## The Conway-Maxwell-Poisson distribution

The negative binomial distribution has traditionally been used as a model for
macroparasite distributions. It has proven an excellent empirical model in
the field [@Pacala1988].  Intuitively, the negative binomial is a mechanistic
sense; it be derived from a continuous mixture of Poisson
distributions with varying means. Thus, it represents a macroparasite
populations colonizing hosts with varying traits or conditions, resulting
in different colonization rates.

However, the negative binomial has a minimum variance/mean ratio of one; it can
not represent underdispersed populations. @Anderson1978 suggested the binomial
distribution to represent these underdispersed populations. This has
the drawback of an absolute maximum number of infections, above which the
probability is zero. In addition, it has the opposite problem of the negative binomial; it can only
represent under-dispersed populations.  Using the negative and positive binomial distributions
means switching between mechanistically different representations of a population.

I compare the fit of the macroparasite distributions generated by the models
above to both the negative binomial and the Conway-Maxwell Poisson (CMP)
distribution [@Conway1962; @Shmueli2005].  The CMP distribution has the form:

$$P(X = x) = \frac{\lambda^x}{(x!)^\nu} \frac{1}{Z(\lambda, \nu)}$$

where

$$Z(\lambda, \nu) = \sum_{j=0}^\infty \frac{\lambda^j}{(j!)^\nu}$$

As in the Poisson distribution, $\lambda$ is an encounter rate. Unlike the 
Poisson
distribution, the $\nu$ term modifies it. When $\nu < 1$, the distribution is 
overdispersed; it has a fatter tail in a Poisson, and its variance is greater 
than its mean. When $\nu > 1$, the distribution is underdispersed. Its tail is 
thinned, and its variance is less than its mean. If $\nu = 1$, the CMP 
distribution is a Poisson distribution. If $\nu = \infty$, the CMP distribution
is the same as a Bernoulli distribution.

The CMP distribution intuitively describes the processes that generate parasite 
distributions. A Poisson process represents the accumulation of parasites, 
while the censorship or extension of the tail, driven by the $\nu$ term,
represents differential accumulation or mortality of parasites according to
host infection number.  

Since the $Z$ term is defined by an infinite series, the CMP density function can
be computationally expensive to calculate. I provide an R package, **cmp**, [REF:TODO]
which contains density, probability, quantile, and random number functions
for the CMP distribution.  It also has functions for fitting this distribution
to data via maximum likelihood or minimum Kullback–Leibler divergence. **cmp** implements the distribution functions using high-performance, parallel C++ code.  It calculates
$Z$ to user-adjustable accuracy, and uses approximations for $Z$ in the areas of
parameter space where they are appropriate [@Shmueli2005], so as to maximize
efficiency.  It also has a C++ API for incorporation into other high-performance
packages. While there have been previous implementations of CMP, they have
not had the combination of accuracy and performance needed for many uses.

# Results

```{r odesims, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results="hide"}
```

```{r odeprocess, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results="hide"}
```

```{r microsims, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results="hide"}
```

```{r microprocess, cache=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results="hide"}
```

Where hosts have low growth rates, parasites invade the host population and go through a period
of high growth, then decline as they and suppress the
host population to very low levels (Figure 1). In the ODE system in the linear case ($p=1$)
the parasite population is underdispersed for an extended transient period.
The variance/mean ratio declines below one, reaching its nadir as the parasite
population declines from its peak and the host population reaches its low values.
The variance/mean ratio eventually equilibriates near one.  At moderate growth
rates, the host population goes through extended dampening cycles before reaching
equilibrium, with repeated parasite population spikes (only one cycle is shown in figures).
In this case, the population passes from underdispersion to overdispersion;
the variance/mean ratio declines modestly below one, then increases
above one in each cycle before reaching equilibrium.  When the hosts have high
growth rates, the dampened cycles are faster and both host and parasite populations
settle at higher values.  In this case, the parasite populations become overdispersed
with the rise in parasite populations and remain overdispersed at equilibrium.

```{r f1, cache=FALSE, fig.cap="Dynamics of a parasite epidemic in the linear ODE model, across different host growth rates $r$. Top (blue): total host population $(N)$, over time.  Middle (red): total parasite population $(P)$ over time. Bottom (green): variance/mean ratio of the distribution of parasites among hosts. Host growth rates vary by column: $r = 0.001$ (left), $r = 0.05$ (middle), $r = 0.2$ (right)."}
<<fig1>>
Figure1
```

The distributions generated by the ODE model are better approximated by the CMP
distribution than the negative binomial at all growth rates.  Figure 2 shows
the relative fit of the two distributions via Kullback–Leibler divergence (KLD). Positive
values indicate a better fit to the CMP than negative binomial. As KLD is log-scale
the y-axis represents the equivalent of a likelihood ratio. In periods of
neither over- nor underdispersion, both distributions fit equally. At all other
times the distributions generated by the ODE model are better approximated
by the CMP distribution.

```{r f2, cache=FALSE, fig.cap="Relative fit of distributions produced by the linear ODE model to negative binomial and CMP distributions.  Positive values indicate better fits for CMP. Host growth rates vary: $r = 0.001$ (left), $r = 0.05$ (middle), $r = 0.2$ (right)."}
<<fig2>>
Figure2
```

Figures 3 and 4 show the system dynamics for the case of $p=2$, or nonlinear
parasite-induced host mortality.  Overall dynamics in this case are similar for
all host growth rates.  However, in this case, underdispersion occurs in at all
growth rates, with weak underdispersion at equilibrium at the highest growth rate.
As would be expected, the CMP distribution describes the distribution of parasites
much better at all growth rates, with the greatest advantage being at low growth
rates, where underdispersion is greatest.


```{r f3, cache=FALSE, fig.cap="Dynamics of a parasite epidemic in the nonlinear ODE model, across different host growth rates $r$. Top (blue): total host population $(N)$, over time.  Middle (red): total parasite population $(P)$ over time. Bottom (green): variance/mean ratio of the distribution of parasites among hosts.  Host growth rates vary by column: $r = 0.001$ (left), $r = 0.05$ (middle), $r = 0.2$ (right)."}
<<fig3>>
Figure3
```

```{r f4, cache=FALSE, fig.cap="Relative fit of distributions produced by the nonlinear ODE model to negative binomial and CMP distributions.  Positive values indicate better fits for CMP. Host growth rates vary: $r = 0.001$ (left), $r = 0.05$ (middle), $r = 0.2$ (right)."}
<<fig4>>
Figure4
```


Figure 5 shows mean dynamics of 100 IBM runs as well as a sampling of individual runs for the linear case. The dynamics of the discrete, stochastic IBM vary in minor ways from the deterministic ODE.  In general, mean host populations
remain higher, and parasite populations are greater at both the peak of the
epidemic and equilibrium. However, the variance/mean ratio does not decline below zero in the IBM in any
of the cases with linear host mortality.  Where the ODE system generates underdispersion,
the IBM is randomly dispersed (variance/mean ratio $\approx$ 1).  Where the ODE generates overdispersion, the IBM
generates stronger overdispersion. This is also reflected in the relative fits of the
distributions (Figure 6). The randomly-dispersed IBM is fit nearly equally well by the CMP and
negatively binomial distributions, while the overdispersed counts are fit better
by the CMP.

```{r f5, cache=FALSE, fig.cap="Dynamics of a parasite epidemic in the linear IBM, across different host growth rates $r$. Mean values of 100 runs shown, with individual runs shown in purple. Top (blue): total host population $(N)$, over time.  Middle (red): total parasite population $(P)$ over time. Bottom (green): variance/mean ratio of the distribution of parasites among hosts.  Orange lines show dynamics of the linear ODE for comparison.  Host growth rates vary by column: $r = 0.001$ (left), $r = 0.05$ (middle), $r = 0.2$ (right)."}
<<fig5>>
Figure5
```

```{r f6, cache=FALSE, fig.cap="Relative fit of distributions produced by the linear IBM model to negative binomial and CMP distributions. Mean value of 100 runs shown, with individual runs shown in purple.  Relative fits for the linear ODE model shown in orange for comparison.  Positive values indicate better fits for CMP. Host growth rates vary: $r = 0.001$ (left), $r = 0.05$ (middle), $r = 0.2$ (right)."}
<<fig6>>
Figure6
```

In the case of nonlinear mortality, the IBM behavior once again largely reflects
that of the ODE (Figure 7).  In this case, however, the underdispersal
in the ODE model is still present, albeit weaker.  The CMP distribution is a
substantially better fit in all cases, though not as strongly as in the ODE case.

```{r f7, cache=FALSE, fig.cap="Dynamics of a parasite epidemic in the nonlinear IBM, across different host growth rates $r$. Mean values of 100 runs shown, with individual runs shown in purple. Top (blue): total host population $(N)$, over time.  Middle (red): total parasite population $(P)$ over time. Bottom (green): variance/mean ratio of the distribution of parasites among hosts.  Orange lines show dynamics of the nonlinear ODE for comparison.  Host growth rates vary by column: $r = 0.001$ (left), $r = 0.05$ (middle), $r = 0.2$ (right)."}
<<fig7>>
Figure7
```

```{r f8, cache=FALSE, fig.cap="Relative fit of distributions produced by the linear IBM model to negative binomial and CMP distributions. Mean value of 100 runs shown, with individual runs shown in purple.  Relative fits for the linear ODE model shown in orange for comparison.  Positive values indicate better fits for CMP. Host growth rates vary: $r = 0.001$ (left), $r = 0.05$ (middle), $r = 0.2$ (right)."}
<<fig8>>
Figure8
```

# Discussion

Underdispersion can occur in classic host-parasite models even in the absence of 
parasite density dependence, though density dependence strengthens this 
effect. It occurs during transient dynamics as parasite population growth drives host
populations down when the time scale of population growth is much slower than
the rate of growth of the parasite. This is a time-delay effect; as high-infection
hosts die from infection, the slow growth of new hosts and their progression
to higher infection levels takes time to rebound,
leading to an overall thinner tail of the distribution during this transient. In the
cases of density-dependent mortality, this underdispersal is enhanced.

This underdispersal effect is weakened or eliminated in discrete IBM models as compared to
ODE models with the same parameterization.  In the IBM, the long tail is truncated, as no values below one are present, while in the ODE the tail may extend much further with below-one values.  This would
intuitively make the distribution more underdispersed.
However, the truncation of the tail also eliminates the outsize effect of these
sub-1 groups from the system, resulting in less overall mortality, weakening
the forces that drive underdispersal.

These models produce distributions more closely matched by the CMP distribution
than the negative binomial in all cases where parasites are non-randomly
distributed.  While this is obvious that would occur for underdispersed
populations, it is somewhat surprising that it did for overdispersed populations.
This highlights the fact that the mechanism producing overdispersal in these
models is different from that commonly-attributed case of variation in host
vulnerability.  Rather, both under- and overdispersal arise from forces of load-
dependent colonization and mortality.  In real systems, parasite distribution
may be driven by a combination of these forces and host variation. Thus,
mixed models, comprised of CMP distributions with varying $\lambda$ terms,
may be appropriate for describing populations when there is sufficient data
and reason to examine multiple forces shaping host-parasite dynamics.

The CMP distribution may be appropriate to other applications in ecology.
Thus far, its use has been limited to a few test cases [@Ridout2004, (clutch size); @Lynch2014 (litter sizes, plant densities, and species richness)]. Mechanistically, CMP is an appropriate
fit for data generated by density-dependent processes or where high counts are
censored. The **cmp** package can be used with other R packages such as **rethinking**
[@rethinking] and **nimble** [@nimble] to build general linear models
and mixed models with the CMP distribution.

# References
