---
title: "Examining Eelgrass Data"
author: "Noam Ross"
date: "February 17, 2015"
output: html_document
---

[Grace Ha](http://www.researchgate.net/profile/Grace_Ha) kindly provided data on seagrass
density and epiphytes to examine for underdispersion with the [cmp package](https://github.com/noamross/cmp),
wich fits distributions to the [Conway-Maxell-Poisson] distribution.  This is
a quick spin through the data.

```{r libraries, message=FALSE}
library(cmp) #get with devtools::install_github('noamross/cmp') if needed
library(dplyr)
library(tidyr)
library(magrittr)
library(ggplot2)
library(knitr)
opts_chunk$set(cache=TRUE)
```

Let's load and take a look at the data.

```{r loaddata, message=FALSE}
ha_data = as.tbl(read.csv('../../data/Bodega-Harbor-Abundance-Data-02182015.csv'))
```

```{r grassplot, message=FALSE, warn=FALSE}
seagrass_counts = ha_data %>%
  filter(site !="Channel") %>%
  mutate(date = as.Date(paste(year, substr(month,1,3), "01"), "%Y %b %d")) %>%
  select(date, site, quadrat, shootdensity) %>%
  distinct(date, site, quadrat)
  

ggplot(seagrass_counts, aes(x = shootdensity)) +
  geom_bar() +
  xlab("Seagrass shoots per quadrat") +
  theme(legend.position="none") +
  ylab("Frequency") +
  facet_grid(date~site)
```

```{r}
seagrass_stats = seagrass_counts %>%
  group_by(date, site) %>%
  summarize(mean_dens = mean(shootdensity), var_dens = var(shootdensity),
            var_mean_ratio = var_dens/mean_dens, samples = n()) %>%
  group_by() %>%
  arrange(site, date)
kable(seagrass_stats, digits = 2)
```

OK, seagrass density is generally highly overdispersed - all the the
variance/mean ratios are over 1.  Let's look at ratios for invertebrates
on leaves:

```{r}
invert_counts = ha_data %>%
  mutate(date = as.Date(paste(year, substr(month,1,3), "01"), "%Y %b %d")) %>%
  gather("invert", "invert_count",  13:67)

invert_stats = invert_counts %>%
  group_by(invert) %>%
  summarize(samples = sum(!is.na(invert_count)),
            nonzeros = sum(na.omit(invert_count) > 0),
            mean_count = mean(invert_count, na.rm=TRUE),
            var_count = var(invert_count, na.rm=TRUE),
            var_mean_ratio = var_count/mean_count) %>%
  group_by() %>%
  arrange(var_mean_ratio)

kable(invert_stats)
```

Most of the invertebrate counts are overdispersed, or have a variance/mean
ratio very close to one.  However, this is when we look at distributions
across all sites and times.  Let's try breaking up by location and date.
In this case, I'll filter to only those that we have more than 1 observation, and
look at only those cases where a species has at least on case where the ratio
is less than one.

```{r}
invert_stats_2 = invert_counts %>%
  group_by(invert, date, site) %>%
    summarize(samples = sum(!is.na(invert_count)),
            nonzeros = sum(na.omit(invert_count) > 0),
            mean_count = mean(invert_count, na.rm=TRUE),
            var_count = var(invert_count, na.rm=TRUE),
            var_mean_ratio = var_count/mean_count) %>%
  group_by(invert, site) %>%
  mutate(mean_ratio = mean(var_mean_ratio, na.rm=TRUE)) %>%
  filter(nonzeros > 1) %>%
  filter(any(var_mean_ratio < 1)) %>%
  group_by() %>%
  arrange(mean_ratio, as.character(invert), site) %>%
  select(-mean_ratio)

kable(invert_stats_2, digits=2)
```

So this table summarizes invertebrate/site/date combinations at the plot level,
showing only those cases where at least one of the invertebrate/site combination
has a variance/mean ratio of less than one.  For the most part, it's hard to see
anything that is consistently underdispersed.

But we don't need to guess!  I'm going to use my `cmp` package to fit a distribution
to each of these

```{r}
invert_fits = invert_counts %>%
  filter(!is.na(invert_count)) %>%
  group_by(invert, date, site) %>%
  filter(any(invert_count > 0)) %>%
  do({
    invert_dist = table(.$invert_count)
    invert_dist = cbind(as.numeric(names(invert_dist)), invert_dist)
    cmp = cmp_fit(invert_dist)
    pois = pois_fit(invert_dist)
    nb = nb_fit(invert_dist)
    data_frame(samples = sum(!is.na(.$invert_count)),
            nonzeros = sum(na.omit(.$invert_count) > 0),
            mean_count = mean(.$invert_count, na.rm=TRUE),
            var_count = var(.$invert_count, na.rm=TRUE),
            var_mean_ratio = var(.$invert_count, na.rm=TRUE)/mean(.$invert_count, na.rm=TRUE),
            cmp_lambda = cmp$lambda, cmp_nu = cmp$nu, cmp_ll = cmp$log.likelihood,
               pois_lambda = pois$lambda, pois_ll = pois$log.likelihood,
               nb_mu = nb$mu, nb_size = nb$size, nb_ll = nb$log.likelihood)
  }) %>%
  group_by() %>%
  arrange(desc(cmp_nu))

kable(invert_fits)
```

