First, a little data mani

```{r setup}
library(dplyr)
library(tidyr)
library(magrittr)
library(ggplot2)
library(compoisson)
```

```{r data_cleanup, eval=FALSE}
dunn = read.csv('data/Dunn-oyster-data.csv')
dunn_sizes = c(Granite=9.381, Marl=9.32, Concrete=9.091, Shell=9.302)
dunn %<>%
  mutate(oyster_count = as.integer(Oyster.Density/dunn_sizes[as.character(Material)]))
names(dunn) = c("river", "site", "material", "sample_date", "oyster_density", "oyster_count")
write.csv(dunn, "data/dunn_oyster_counts.csv")
```

```{r gatherdata}
dunn = read.csv('../../data/dunn_oyster_counts.csv')
freqdata = dunn %>%
  group_by(material) %>%
  do({freq = table(.$oyster_count)
      count = as.integer(names(freq))
      data_frame(freq, count)}) %>%
  arrange(material, count)

```{r plotfreqs}
ggplot(freqdata, aes(x=count, y=freq)) +
  geom_bar(stat="identity", position="identity") +
  facet_wrap(~material)
```

```{r fitdists}
fits = freqdata %>%
  group_by(material) %>%
  do({
    N = sum(.$count * .$freq)
    com = com.fit(cbind(.$count, .$freq))
    pois = pois.fit(cbind(.$count, .$freq))
    nb = nb.fit(cbind(.$count, .$freq))
    data_frame(N, com_lambda = com$lambda, com_nu = com$nu, com_ll = com$log.likelihood,
               pois_lambda = pois$lambda, pois_ll = pois$log.likelihood,
               nb_mu = nb$mu, nb_size = nb$size, nb_ll = nb$log.likelihood)
  })
```

```{r}
max_val = max(freqdata$count)

sims = fits %>%
  group_by(material) %>%
  do(data_frame(count = 0:max_val,
                sim_com = dcom(0:max_val, .$com_lambda, .$com_nu),
                sim_pois = dpois(0:max_val, .$pois_lambda),
                sim_nb = dnbinom(0:max_val, mu = .$nb_mu, size=.$nb_size)))
sims = full_join(sims, freqdata, by=c('material', 'count')) %>%
  mutate(freq = replace(freq, is.na(freq), 0)/sum(freq, na.rm=TRUE))

```

```{r plotsims}
sims2 = sims %>%
  gather(origin, frequency, sim_com, sim_pois, sim_nb, freq)
ggplot(sims) +
  geom_bar(mapping = aes(x = count, y = frequency),
           data = filter(sims2, origin=="freq"),
           stat="identity", position="identity") +
  geom_step(mapping = aes(x = count - 0.5, y = frequency, col=origin),
            data = filter(sims2, origin!="freq")) + 
  facet_wrap(~material, ncol=1)
```